sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-09-09 doc/paper-jss/paper-jss.Rmd
Edits

@@

@@

@@

```{r pcod-predict-maps, fig.width=11, fig.asp=0.3, out.width="6in", fig.cap="TODO"}
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position = "bottom") +
    ggtitle(title)
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
```

@@

```{r plot-map, echo=FALSE}
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(vars(year)) +
    coord_fixed() +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
```

@@

```{r dog-wvci-pred, fig.asp = 0.5, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
```

@@

The spatially varying coefficient effect at any point in space is the
combination of the main effect and spatial random effect for `nao`. Here we
illustrate deriving uncertainty on their combination by simulating from the
parameter precision matrix assuming the joint parameter covariance matrix in
multivariate normal. We do this by specifying a non-null number of simulation
draws to `nsim` in `predict.sdmTMB()`. By default, `nsim > 0` will return
a matrix of draws from the overall prediction. Here, we we instead specify that
we want to return draws from the `zeta_s` random field, which is the spatially
varying random field. This returns a matrix where each row matches a row of
`newdata` and each column is a simulation draw. We then use the function
`spread_sims()` to draw 200 simulations from the parameters themselves, combine
the main effect draws (`sims$nao`) with the random field values `zeta_s`, and
calculate the median for each row of data so we can plot them.

@@

```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the lower values are near 1 implying no effect, whereas in the southeast the effect becomes positive."}
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
```

@@

```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
# tidy(fit, conf.int = TRUE)
```

```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
# tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

>>>>

TODO ADD EQUATIONS

@@

TODO: BEFORE OR AFTER EXAMPLES

@@

We can also use `tidy()` to get a data frame of our random field parameters,
where `sigma_O` is $\sigma_O$ in the above equations: the standard deviation of
the random field.

```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="TODO"}
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position= "bottom") +
    ggtitle(title) +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
```

@@

```{r plot-map, echo=FALSE}
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(vars(year)) +
    coord_fixed() +
    theme(legend.position= "bottom") +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
```

@@

```{r dog-wvci-pred, fig.asp = 0.7, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
```

@@

The spatially varying coefficient effect at any point in space is the
combination of the main effect and spatial random effect for `nao`. We
illustrate deriving uncertainty on their combination by simulating from the
parameter precision matrix assuming the joint parameter covariance matrix in
multivariate normal. We do this by specifying a non-null number of simulation
draws to `nsim` in `predict.sdmTMB()`. By default, `nsim > 0` will return
a matrix of draws from the overall prediction. Here, we instead specify that we
want to return draws from the `zeta_s` random field, which is the spatially
varying random field. This returns a matrix where each row matches a row of
`newdata` and each column is a simulation draw. We then use the function
`spread_sims()` to draw 200 simulations from the parameters themselves, combine
the main effect draws (`sims$nao`) with the random field values `zeta_s`, and
calculate the median for each row of data so we can plot them.

@@

```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5}
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
```

@@
